---
title: "GenEst Command Line Example"
author: "Juniper L. Simonis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(devtools)
library(matrixStats)
library(survival)
library(mvtnorm)
library(matrixStats)
library(gsl)
library(GenEst)
```

This vignette walks through a full example of GenEst at the command line. 


## Vignette Info

Created with GenEst version 0.0.2.1

## Preliminaries

### Read in data

Read in the Searcher Efficiency, Carcass Persistence, Search Schedule, 
and Carcass observation datasets. 


```{r, results = "hide"}
SEdataIn <- read.csv("data/ExampleSearcherEfficiency.csv")
CPdataIn <- read.csv("data/ExampleCarcassPersistence.csv")
SSdataIn <- read.csv("data/ExampleSearchSchedule.csv")
COdataIn <- read.csv("data/ExampleCarcassObservations.csv") 
```

### General Inputs

Number of replicate draws used to estimate distributions

```{r, results = "hide"}
Niterations <- 1000
```

## Searcher Efficiency

### Preliminaries

Select predictors, observation columns, and size class column

```{r, results = "hide"}
SEvars <- c("Season", "HabitatType")
SEobscols <- 8:11
SEsizeclasscol <- "Size"
```
  
Declare if you will fix k or not and if so, the value.

```{r, results = "hide"}
fixKchoice <- FALSE
fixKvalchoice <- NULL
```

### Analyses

Run all searcher efficiency models across all size classes

```{r, results = "hide"}
SEmods <- se_model_set_across_sizes_fit(data = SEdataIn, predictors = SEvars, 
                                        observation_columns = SEobscols, 
                                        size_class_column = SEsizeclasscol,
                                        init_k_value = 0.7, fix_k = fixKchoice, 
                                        fix_k_value = fixKvalchoice)
```

Create a theta for each cell within each each model in each size class
(dimension: [Niterations, 2, Ncells(SE), Nmodels(SE), Nsizeclasses])

```{r, results = "hide"}
thetaSE <- se_theta_create(data = SEdataIn, predictors = SEvars,
                           size_class_column = SEsizeclasscol,
                           model_fits = SEmods, replicates = Niterations,
                           fix_k = fixKchoice, fix_k_value = fixKvalchoice)
```
	
### Results
	
Create the AICc table for all sizes

```{r, result = "hide"}
SEtable <- se_aicc_table_create(models = SEmods)
```

AICc table for Small size class

```{r, result = "asis"}
knitr::kable(SEtable$S)
```

Plot the results

```{r, fig.width = 8, fig.height = 8}
create_se_figure(data = SEdataIn, predictors = SEvars, theta = thetaSE, 
                 observation_columns = SEobscols, replicates = Niterations, 
                 size_class_column = SEsizeclasscol, r = 4, j = 1, 
				 cellwise = 25)
```


## Carcass Persistence

### Preliminaries

Select predictors, observation columns, and size class column

```{r, results = "hide"}
CPvars <- NULL
CPsizeclasscol <- "Size"
CPltp <- "LastPresentDecimalDays"
CPfta <- "FirstAbsentDecimalDays"
```
	  
### Analyses

Run the estimator for each of the possible models for each size class

```{r, results = "hide"}
CPmods <- cp_model_set_across_sizes_fit(data = CPdataIn, predictors = CPvars,
                                        size_class_column = CPsizeclasscol, 
                                        last_time_present_column = CPltp, 
                                        first_time_absent_column = CPfta)
```

Create a theta for each cell within each each model in each size class
(dimension: [Niterations, 2, Ncells(CP), Nmodels(CP), Nsizeclasses])

```{r, results = "hide"}
thetaCP <- cp_theta_create(data = CPdataIn, predictors = CPvars,
                           size_class_column = CPsizeclasscol,
                           model_fits = CPmods, replicates = Niterations)
```
										  
### Results
										  
Create the AICc table for all sizes

```{r, result = "hide"}
CPtable <- cp_aicc_table_create(models = CPmods)
```

AICc table for Small size class

```{r, result = "asis"}
knitr::kable(CPtable$S)
```

Plot the results

```{r, fig.width = 8, fig.height = 8}
create_cp_figure(models = CPmods, data = CPdataIn, predictors = CPvars,
                 theta = thetaCP, time_unit = "days",
                 size_class_column = CPsizeclasscol,
                 last_time_present_column = CPltp, 
                 first_time_absent_column = CPfta, r = 1, 
                 model_complexity = "~ 1", 
                 distribution_choice = "exponential")
```



## Detection Probability

### Preliminaries

Select the models to use for each size class

```{r, results = "hide"}
CPmodstouse <- c(1, 1, 1, 1)
SEmodstouse <- c(1, 1, 1, 1)
```
	  
### Analyses

```{r, results = "hide"}
garray <- estimate_g_across_sizes(cp_data = CPdataIn, se_data = SEdataIn, 
                                  ss_data = SSdataIn, replicates = Niterations, 
                                  cp_predictors = CPvars, cp_theta = thetaCP,
								  se_predictors = SEvars, se_theta = thetaSE,  
								  cp_models = CPmods,
                                  se_models_to_use = SEmodstouse, 
                                  cp_models_to_use = CPmodstouse)
```

## Fatalities

### Analyses

```{r, results = "hide"}
Mhatarray <- estimate_mhat(co_data = COdataIn, ss_data = SSdataIn, 
                           size_class_column = "Size", split_column = "Split", 
                           unit_column = "Unit", df_column = "DateFound",
                           replicates = Niterations, cp_predictors = CPvars, 
                           se_predictors = SEvars, cp_data = CPdataIn, 
                           se_data = SEdataIn, garray = garray) 
```

Condense Mhat to split categories (dimension: [Niterations, Nsplitcats])
    
```{r, results = "hide"}	
Mhatsc <- condense_mhat(Mhatarray)
```

### Results
	  
Produce Mhat table (allows for expansion to whole facility )

```{r, results = "hide"}
Mhattab <- create_mhat_table(condensed_mhat = Mhatsc, 
                             fraction_area_sampled = 0.85, 
                             confidence_level = 0.9)
```

```{r, result = "asis"}
knitr::kable(Mhattab)
```

Plot Mhat for each split

```{r, fig.width = 8, fig.height = 8}
par(mfrow = c(1, 2))
l <- 1
create_mhat_figure(condensed_mhat_split = Mhatsc[,l], 
                   split_category_name = colnames(Mhatsc)[l], 
                   fraction_area_sampled = 0.85)
l <- 2
create_mhat_figure(condensed_mhat_split = Mhatsc[,l], 
                   split_category_name = colnames(Mhatsc)[l], 
                   fraction_area_sampled = 0.85)
```

