---
title: "GenEst 101--Tutorial with Examples"
author: "Dan Dalthorp"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GenEst 101--Tutorial with Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(GenEst)
vers <- packageVersion("GenEst")
today <- Sys.Date()
```
# Introduction: Tutorial with Examples
This tutorial provides an introduction to the array of command line tools
**GenEst** provides for estimating carcass arrivals and detection
probabilities. The approach is to walk through analyses of realistic but 
fictitious data sets representing studies of bird and bat mortality at a
wind power facility.

The tutorial was constructed using **GenEst** version `r vers` on `r today`.

## Example 1: Estimating Bat Mortality from Searches on Roads and Pads

Searcher efficiency and carcass persistence would be expected to vary with
carcass size (sparrow, eagle, bat), ground characteristics (road & pad,
cleared field, vegetation), season, etc. In this first example, we limit the
basic mortality estimation to one carcass size (`bat`) and one ground 
visibility class (`RP` = road and pad). Once the initial estimates have been 
calculated, the total mortality can be parsed into any number of "splits", 
estimating mortality by species, turbine, visibility class, season, search 
interval, month, etc. or by combinations of splitting covariates, like species 
by season, turbine by search interval, etc.

Load the full data set into R:
```{r}
data(wind_RPbat)
names(wind_RPbat)
```
`wind_RPbat` is a list of data frames that store results for searcher 
efficiency (`SE`) and carcass persistence trials (`CP`), search schedules for 
all turbines (`SS`), the density weighted proportion (`DWP`) of area searched
at each turbine (i.e., the fraction of carcasses expected to fall in the 
search plots), and a carcass observation (`CO`) data. To streamline the 
notation, we will extract the data from the `wind_RPbat` list into its 
components:
```{r}
SE_data <- wind_RPbat$SE
CP_data <- wind_RPbat$CP
SS_data <- wind_RPbat$SS
DWP_data <- wind_RPbat$DWP
CO_data <- wind_RPbat$CO
```

```{r, include = FALSE}
daterange <- range(SS_data$SearchDate)
seasons <- paste(unique(SS_data$Season), collapse = ', ')
```

### Searcher Efficiency (`SE`)
Searcher efficiency trials were conducted on roads and pads, with a total of
`r  dim(SE_data)[1]` fresh carcasses placed in the field over the course of 
the entire monitoring period, evenly divided among seasons (`r seasons`). 
Carcasses that were later discovered by search teams during the course of 
normal carcass surveys were removed from the field. Carcasses were left in 
the field for up to 5 searches after carcass placement.

Results of the SE field trials are stored in the `SE_data` data frame:
```{r}
head(SE_data)
```
Columns `s1, s2,` ... show the fate of carcass `pkID` on the 1st, 2nd, ...
searches after the carcass was placed. A 1 indicates that the carcass was
discovered, a 0 indicates that the carcass was present but not discovered, and
NA indicates that the carcass was not present (whether it was discovered in a
previous search or removed by scavengers).

### Carcass Persistence (`CP`)
Carcass persistence trials were conducted on roads and pads, with a total of
`r  dim(CP_data)[1]` fresh carcasses placed in the field over the course of 
the entire monitoring period, evenly divided among seasons (`r seasons`). 
Carcasses were checked (approximately) 1, 2, 3, 4, 7, 10, 14, 21, and 28 
days after placement in the field (exact times were entered as decimal 
fractions of days after placement).

Results of the SE field trials are stored in the `CP_data` data frame:
```{r}
head(CP_data)
```

Exact scavenging times are not known, but a carcass that was present at one
check and absent at the next check is assumed to have been scavenged at some
point in the interval. The left endpoint of the interval is entered as `Left`
and the right endpoint as `Right`. For carcasses that had not been scavenged
by the end of the study, `Left` is the time of the last check and `Right` 
is `Inf`. For carcasses whose scavenging time is known exactly (e.g., the 
scavenging was recorded by camera), `Left = Right`.

The `Season` column gives the season at the time the carcass was placed in the
field.

### Search Schedules (`SS`)
Carcass searches were conducted on roads and pads within a 120 m radius from 
all `r length(unique(SS_data$SearchDate))` turbines at a fictitious wind power
facility. Monitoring began on `r daterange[1]` and continued through
`r daterange[2]`. Searches spanned `r length(unique(SS_data$Season))` seasons:
`r seasons`. Search intervals varied by turbine and by time of year, ranging 
from daily searches at some turbines in the fall and searches once every 12 
days in the spring at some other turbines. Search schedules for all turbines 
are stored in `SS_data`, which is a data frame with a column for search dates 
(including all dates that any turbine was searched); a column of 0s and 1s for 
each turbine, indication whether it was searched on the given date; and zero 
or more optional columns giving additional information about the date (e.g.,
season).
```{r}
head(SS_data[, 1:10])
print("NOTE: there are 100 turbine columns altogether (t1, ..., t100)")
```

### Density Weighted Proportion (`DWP`)
```{r}
head(DWP_data)
```

### Carcass Observations (`CO`)
```{r}
head(CO_data)
```

```{r}
SE_model <- pkm(p ~ 1, k ~ 1, data = SE_data)
CP_model <- cpm(l ~ 1, s ~ 1, data = CP_data, left = "Left", right = "Right")


nsim <- 100
Mest <- estM(
  nsim = nsim, data_CO = CO_data, data_SS = SS_data,
  data_DWP = DWP_data, frac = 1, model_SE = SE_model, model_CP = CP_model,
  seed_SE = NULL, seed_CP = NULL, seed_g = NULL, seed_M = NULL, kFill = NULL,
  unitCol = "Turbine", dateFoundCol = "DateFound",
  datesSearchedCol = "SearchDate", DWPCol = "bat"
)
summary(Mest)
plot(Mest)

M_season <- calcSplits(
  M = Mest$M, Aj = Mest$Aj,
  split_SS = "Season", data_SS = SS_data,
  split_CO = NULL,  data_CO = CO_data
)

summary(M_season)
plot(M_season)

SSdat <- SS(SS_data)
M_week <- calcSplits(
  M = Mest$M, Aj = Mest$Aj,
  split_time = seq(0, max(SSdat$days), by = 7), 
  data_SS = SSdat,
  data_CO = CO_data
)
plot(M_week, rate = T)

M_unit <- calcSplits(
  M = Mest$M, Aj = Mest$Aj,
  split_CO = "Turbine", 
  data_CO = CO_data, 
  data_SS = SS_data
)
plot(M_unit, rate = F)
summary(M_unit)
summary(M_unit, CL = 0.8)

M_unit_and_species <- calcSplits(
  M = Mest$M, Aj = Mest$Aj,
  split_CO = c("Turbine", "Species"),
  data_CO = CO_data,
  data_SS = SS_data
)
plot(M_unit_and_species, rate = F)


M_by_week_and_speciesgroup <- calcSplits(
  M = Mest$M, Aj = Mest$Aj,
  split_CO = "SpeciesGroup",
  data_CO = CO_data,
  split_time = seq(0, max(SSdat$days), by = 7),
  data_SS = SSdat
)
plot(M_by_week_and_speciesgroup, rate = T)
```
