---
title: "Tutorial: GenEst command line tools"
author: "Dan Dalthorp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(GenEst)
vers <- packageVersion("GenEst")
today <- Sys.Date()
```

This tutorial provides an introduction to the array of command line tools
**GenEst** provides for estimating carcass arrivals and detection probabilities.
The approach is to walk through analyses of realistic but fictitious data sets
representing studies of bird and bat mortality at a wind power facility.

The tutorial was constructed using **GenEst** version `r vers` on `r today`.

```{r}

data(wind_RP)
batind <- wind_RP[["SE"]]$Size == "bat"
SE_data <- wind_RP[["SE"]][batind,]

batind <- which(wind_RP[["CP"]]$Size == "bat")
CP_data <- wind_RP[["CP"]][batind,]

DWP_data <- wind_RP[["DWP"]][ , c("Turbine", "DWP_bat")]

batind <- which(wind_RP[["CO"]]$Size == "bat")
CO_data <- wind_RP[["CO"]][batind, ]

SS_data <- wind_RP[["SS"]]

pkModel <- pkm(p ~ 1, k ~ 1, data = SE_data)
cpModel <- cpm(l ~ 1, s ~ 1, data = CP_data, left = "Left", right = "Right")

tst <- proc.time()
nsim <- 1000
gAj <- rghat(n = nsim,
  data_CO = CO_data,
  data_SS = SS_data,
  model_SE = pkModel,
  model_CP = cpModel,
  unitCol = "Turbine",
  dateFoundCol = "DateFound",
  dateSearchedCol = "SearchDate",
  removeCleanout = TRUE
)
ghat <- gAj$ghat
Aj <- gAj$Aj
proc.time() - tst # 4 minutes!

DWP <- DWPbyCarcass( # doesn't work
  data_DWP = DWP_data,
  data_carc = CO_data,
  unitCol = "Turbine",
  sizeclassCol = NULL,
  removeCleanout = TRUE,
  data_SS = SS_data,
  dateFoundCol = "DateFound",
  dateSearchedCol = "SearchDate"
)
DWP <- DWP_data$DWP_bat[match(rownames(Aj), DWP_data$Turbine)]
Mhat <- rMhat(n = 1, ghat = ghat, DWP, seed = 12)

SSdat <- SS(SS_data)

M_by_season <- calcSplits(
  M = Mhat, Aj = Aj,
  split_SS = "Season",
  data_SS = SS(SS_data)
)

plot(M_by_season)
M_by_week <- calcSplits(M = Mhat, Aj = Aj,
  split_time = seq(0, max(SSdat$days), by = 7), data_SS = SSdat)
plot(M_by_week, rate = T)
M_by_unit <- calcSplits(M = Mhat, Aj = Aj, split_CO = "Turbine", data_CO = CO_data)
plot(M_by_unit, rate = F)
summary(M_by_unit)
summary(M_by_unit, CL = 0.9)
```
