---
title: "GenEst Command Line Example"
author: "Juniper L. Simonis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(GenEst)
vers <- packageVersion("GenEst")
today <- Sys.Date()
```

This vignette walks through an example of **GenEst** at the command line and
was constructed using **GenEst** version `r vers` on `r today`.  


## Installation

To obtain the most recent version of **GenEst**, use the **devtools** package 
to download it from GitHub:

```{r, results = "hide", message = FALSE, warning = FALSE}
install.packages("devtools")
devtools::install_github("ddalthorp/GenEst")
```

## Data

For this vignette, we will be using a completely mock dataset provided with 
the **GenEst** package, which contains Searcher Efficiency, Carcass
Persistence, Search Schedule, Density Weighted Proportion, and Carcass 
Observation Data. 

```{r}
data(mockData)
names(mockData)
```

To see the raw `.csv` versions of the data files, find the **GenEst** 
package folder on your computer (it should be located with all of your other
R packages) and then browse to the `inst/extdata/mockData` subfolder.


## Searcher Efficiency

### Single Searcher Efficiency Model

The central function for searcher efficiency analyses is `pkm`, which conducts
a singular searcher efficiency analysis (*i.e.*, a singular set of `p` and 
`k` formulae and a singular size classification of carcasses). As a first 
example, we will ignore the size category and use intercept-only models for 
both `p` and `k`:

```{r}
dataSE <- mockData$SearcherEfficiencyData
pkModel <- pkm(formula_p = p ~ 1, formula_k = k ~ 1, data = dataSE)
```

Here, we have taken advantage of `pkm`'s default behavior of selecting 
observation columns (see `?pkm` for details). 

```{r}
head(dataSE)
```

If we wanted to explicitly control the observations, we would use the `obsCol` 
argument:

```{r}
pkModel <- pkm(formula_p = p ~ 1, formula_k = k ~ 1, data = dataSE, 
             obsCol = c("Search1", "Search2", "Search3", "Search4")
           )
```

Note that the search observations must be entered in order such that no 
carcasses have non-detected observations (*i.e.*, `0`) after detected
observations (*i.e.*, `1`). Further, no carcasses can be detected more than 
once.

If successfully fit, a `pkm` model output contains a number of elements, 
some printed automatically:

```{r}
pkModel
```

and others available upon request (using `$` as `pkm` is an S3 class):

```{r}
names(pkModel)
pkModel$cells
```

The `plot` function has been defined for `pkm` objects, such that one can
simply run

```{r, fig.show = "hold", fig.width = 7, fig.height = 7}
plot(pkModel)
```

to visualize the model's output. 

You can generate random draws of the `p` and `k` parameters for each cell 
grouping (in `pkModel` there are no predictors, so there is one cell grouping
called "all") using the `rpk` function which, like other `r*` functions in 
**R** (*i.e.*, `rnorm`) takes the number of random draws (`n`) as the first
argument:

```{r}
rpk(n = 10, pkModel)
```

You can complicate the `p` and `k` formulae independently

```{r}
pkm(formula_p = p ~ Visibility, formula_k = k ~ HabitatType, data = dataSE,
  obsCol = c("Search1", "Search2", "Search3", "Search4")
)
```

And you can fix `k` at a nominal value between 0 and 1 (inclusive) using the
`kFixed` argument

```{r}
pkm(formula_p = p ~ Visibility, kFixed = 0.7, data = dataSE,
  obsCol = c("Search1", "Search2", "Search3", "Search4")
)
```

### Set of Searcher Efficiency Models

The function `pkmSet` provides a simple wrapper on `pkm` that facilitates
analyzing a set of `pkm` models defined as all allowable models simpler than, 
and including, the provided model formulae (where "allowable" means that 
any interaction terms have all component terms included in the model). 

Consider the following model set analysis, where visibility and habitat type
are included in the `p` formula but only habitat type is in the `k` formula,
which generates a set of 10 models:

```{r}
pkmModSet <- pkmSet(formula_p = p ~ Visibility*HabitatType, 
               formula_k = k ~ HabitatType, data = dataSE,
               obsCol = c("Search1", "Search2", "Search3", "Search4")
             )
names(pkmModSet)
```

The `plot` function is defined for the `pkmSet` class, and by default, 
creates a new plot window on command for each sub-model. If we want to only
plot a specific single (or subset) of models from the full set, we can utilize
the `specificModel` argument:

```{r, fig.show = "hold", fig.width = 7, fig.height = 7}
plot(pkmModSet, specificModel = "p ~ Visibility + HabitatType; k ~ 1")
```

The resulting model outputs can be compared in an AICc table 

```{r}
pkmSetAICcTab(pkmModSet)
```


### Multiple Sizes of Animals and Sets of Searcher Efficiency Models

Often, carcasses are grouped in multiple size classes, and we are interested 
in analyzing a set of models separately for each size class. To do so, we
use the `pkmSetSize` function, which is a wrapper on `pkmSet` that runs
the subfunction for each unique size class in the column identified by the
`sizeclassCol` argument:

```{r}
pkmModSetSize <- pkmSetSize(formula_p = p ~ Visibility*HabitatType, 
                   formula_k = k ~ HabitatType, data = dataSE,
                   obsCol = c("Search1", "Search2", "Search3", "Search4"),
                   sizeclassCol = "Size"
                 )
```

The output from `pkmSetSize` is a list where each element corresponds to a
different unique size class, and contains the associated `pkmSet`object, which
itself is a list of `pkm` outputs:
 
```{r}
names(pkmModSetSize)
names(pkmModSetSize[[1]])
```



## Carcass Persistence

### Single Carcass Persistence Model

The central function for carcass persistence analyses is `cpm`, which conducts
a singular carcass persistence analysis (*i.e.*, a singular set of `l` and 
`s` formulae and a singular size classification of carcasses). Note that
we use `l` and `s` to reference `location` and `scale` as the parameters for
survival models, following `survreg`, however we also provide an alternative
parameterization ("`ab`"). As a first example, we will ignore the size 
category, use intercept-only models for both `l` and `s`, and use the 
Weibull distribution:

```{r}
dataCP <- mockData$CarcassPersistenceData
cpModel <- cpm(formula_l = l ~ 1, formula_s = s ~ 1, data = dataCP,
             left = "LastPresentDecimalDays", 
             right = "FirstAbsentDecimalDays", dist = "weibull"
           )
```


If successfully fit, a `cpm` model output contains a number of elements, 
some printed automatically:

```{r}
cpModel
```

and others available upon request (using `$` as `cpm` is an S3 class):

```{r}
names(cpModel)
cpModel$cells
```

The `plot` function has been defined for `cpm` objects, such that one can
simply run

```{r, fig.show = "hold", fig.width = 7, fig.height = 7}
plot(cpModel)
```

to visualize the model's output. 


You can generate random draws of the `l` and `s` (or `a` and `b`) parameters
for each cell grouping (in `cpModel` there are no predictors, so there is one
cell grouping called "all") using the `cpk` function which, like other `r*` 
functions in **R** (*i.e.*, `rnorm`) takes the number of random draws (`n`)
as the first argument:

```{r}
rcp(n = 10, cpModel)
rcp(n = 10, cpModel, type = "ppersist")
```

You can complicate the `l` and `s` formulae independently

```{r}
cpm(formula_l = l ~ Visibility*GroundCover, formula_s = s ~ 1, data = dataCP,
  left = "LastPresentDecimalDays", right = "FirstAbsentDecimalDays", 
  dist = "weibull"
)
```

Given that the exponential only has one parameter (`l`, location), a model 
for scale (`formula_s`) is not required:

```{r}
cpModExp <- cpm(formula_l = l ~ Visibility*GroundCover, data = dataCP,
              left = "LastPresentDecimalDays", 
              right = "FirstAbsentDecimalDays", dist = "exponential"
            )
```


### Set of Carcass Persistence Models

The function `cpmSet` provides a simple wrapper on `cpm` that facilitates
analyzing a set of `cpm` models defined as all allowable models simpler than, 
and including, the provided model formulae (where "allowable" means that 
any interaction terms have all component terms included in the model). 

In addition, `cpmSet` includes the four base distributions (exponential, 
weibull, lognormal, loglogistic) and crosses them with the predictor models,
but this can be restricted using the `dist` argument.

Consider the following model set analysis, where visibility and season are
included in the `l` formula but only visibility is in the `s` formula, and
all only the exponential and lognormal distributions are included, which 
generates a set of 15 models:

```{r}
cpmModSet <- cpmSet(formula_l = l ~ Visibility*Season, 
               formula_s = s ~ Visibility, data = dataCP,
               left = "LastPresentDecimalDays", 
               right = "FirstAbsentDecimalDays", 
               dist = c("exponential", "lognormal")
             )
names(cpmModSet)
```

The resulting model outputs can be compared in an AICc table 

```{r}
cpmSetAICcTab(cpmModSet)
```

The `plot` function is defined for the `cpmSet` class, and by default, 
creates a new plot window on command for each sub-model. If we want to only
plot a specific single (or subset) of models from the full set, we can utilize
the `specificModel` argument:

```{r, fig.show = "hold", fig.width = 7, fig.height = 7}
plot(cpmModSet, 
  specificModel = "dist: lognormal; l ~ Visibility * Season; s ~ Visibility")
```



### Multiple Sizes of Animals and Sets of Carcass Persistence Models

Often, carcasses are grouped in multiple size classes, and we are interested 
in analyzing a set of models separately for each size class. To do so, we
use the `cpmSetSize` function, which is a wrapper on `cpmSet` that runs
the subfunction for each unique size class in the column identified by the
`sizeclassCol` argument:

```{r}
cpmModSetSize <- cpmSetSize(formula_l = l ~ Visibility*Season, 
                   formula_s = s ~ Visibility, data = dataCP,
                   left = "LastPresentDecimalDays", 
                   right = "FirstAbsentDecimalDays", 
                   dist = c("exponential", "lognormal"),
                   sizeclassCol = "Size"
                 )
```

The output from `cpmSetSize` is a list where each element corresponds to a
different unique size class, and contains the associated `cpmSet`object, which
itself is a list of `cpm` outputs:
 
```{r}
names(cpmModSetSize)
names(cpmModSetSize[[1]])
```
